---
title: "AlphaEarth/SOC — Follow-on Analysis"
subtitle: "Multi-target, S2 baseline, KSSL validation"
author: "Luke Robinson"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5, dpi = 300,
  fig.path = "../figures/", dev = "png"
)
set.seed(42)

suppressPackageStartupMessages({
  library(tidyverse)
  library(boot)
  library(ranger)
  library(car)
  library(patchwork)
  library(scales)
})

dir.create("../figures", showWarnings = FALSE, recursive = TRUE)

ae_cols <- sprintf("A%02d", 0:63)

# Spatial-block CV identical to the original paper.
make_spatial_folds <- function(df, k = 5, seed = 42) {
  set.seed(seed)
  df$block <- paste0(round(df$lon), "_", round(df$lat))
  ub <- unique(df$block)
  fa <- sample(seq_len(k), length(ub), replace = TRUE)
  df$fold <- fa[match(df$block, ub)]
  df
}

# Boot-CI helper: mean R^2 across folds.
boot_mean_r2 <- function(r2_vec, B = 10000) {
  b <- boot::boot(r2_vec, function(x, idx) mean(x[idx]), R = B)
  ci <- boot::boot.ci(b, type = "bca")$bca[4:5]
  list(mean = mean(r2_vec), lo = ci[1], hi = ci[2])
}

# Single-target spatial-CV RF; returns per-fold R2/RMSE plus oof predictions.
rf_spatial_cv <- function(df, target, features, num_trees = 500, seed = 42) {
  form <- as.formula(paste(target, "~", paste(features, collapse = "+")))
  out <- map_dfr(sort(unique(df$fold)), function(k) {
    tr <- df %>% filter(fold != k)
    te <- df %>% filter(fold == k)
    m  <- ranger(form, data = tr, num.trees = num_trees, seed = seed)
    pr <- predict(m, data = te)$predictions
    tibble(fold = k, n = nrow(te),
           rmse = sqrt(mean((pr - te[[target]])^2)),
           r2   = cor(pr, te[[target]])^2,
           pred = list(pr), obs = list(te[[target]]),
           lon = list(te$lon), lat = list(te$lat))
  })
  out
}
```

# Phase A — Sanity check

```{r load-unified}
df <- read.csv("../data/soil_unified.csv")
cat("rows:", nrow(df), " cols:", ncol(df), "\n")
df$log_soc <- log1p(df$soc)
df$ph <- df$ph_x10 / 10
# Logit transform sand/clay (bounded [0,100])
logit_pct <- function(x) log((x/100 + 1e-3) / (1 - x/100 + 1e-3))
df$sand_logit <- logit_pct(df$sand_pct)
df$clay_logit <- logit_pct(df$clay_pct)
df$land_cover <- as.character(df$land_cover)

# Pre-flight ranges
cat("SOC range:", range(df$soc), "\n")
cat("pH range:",  range(df$ph), "\n")
cat("sand range:", range(df$sand_pct), "\n")
cat("clay range:", range(df$clay_pct), "\n")
cat("BD range:",   range(df$bd), "\n")
cat("AE range:",   range(unlist(df[, ae_cols])), "\n")
```

```{r sanity-rf}
df <- make_spatial_folds(df)
rf_check <- rf_spatial_cv(df, "log_soc", ae_cols)
cat("AE -> log-SOC paper-recipe R2 per fold:\n"); print(rf_check %>% select(fold, n, rmse, r2))
ci <- boot_mean_r2(rf_check$r2)
cat(sprintf("Mean R2 = %.3f  (BCa 95%% CI: [%.3f, %.3f])\n", ci$mean, ci$lo, ci$hi))

stopifnot(ci$mean >= 0.70, ci$mean <= 0.78)  # sanity gate
```

# Phase B — Multi-target analysis (Direction 1)

```{r multi-targets}
targets <- list(
  log_soc    = list(label = "log(1 + SOC)",       raw = "soc"),
  ph         = list(label = "pH (H2O)",            raw = "ph"),
  sand_logit = list(label = "logit(sand %)",       raw = "sand_pct"),
  clay_logit = list(label = "logit(clay %)",       raw = "clay_pct"),
  bd         = list(label = "Bulk density (kg/m3)",raw = "bd")
)
target_names <- names(targets)
```

## Distributions and skewness

```{r multi-dist, fig.cap="Distributions of the five soil targets after the chosen transforms."}
df_long <- df %>%
  select(log_soc, ph, sand_logit, clay_logit, bd) %>%
  pivot_longer(everything(), names_to = "target", values_to = "value")
ggplot(df_long, aes(value)) +
  geom_histogram(bins = 40, fill = "#5b8def", color = "white") +
  facet_wrap(~ target, scales = "free", ncol = 3) +
  theme_minimal()
```

## Per-target Welch's t (Grasslands vs. Croplands)

```{r ht-multi}
gg <- df %>% filter(land_cover %in% c("10","12"))
ht_results <- map_dfr(target_names, function(t) {
  res <- t.test(reformulate("land_cover", t), data = gg)
  g1 <- gg[[t]][gg$land_cover == "10"]
  g2 <- gg[[t]][gg$land_cover == "12"]
  psd <- sqrt(((length(g1)-1)*var(g1) + (length(g2)-1)*var(g2)) /
              (length(g1)+length(g2)-2))
  d   <- (mean(g1) - mean(g2)) / psd
  tibble(target = t, t_stat = res$statistic, df = res$parameter,
         p_value = res$p.value, ci_lo = res$conf.int[1], ci_hi = res$conf.int[2],
         cohens_d = d)
})
ht_results
```

## 5 × 64 univariate correlation barplot

```{r multi-corr-bar, fig.cap="Pearson correlation between each AlphaEarth dimension and each soil target.", fig.height=8}
corr_long <- map_dfr(target_names, function(t) {
  rs <- sapply(ae_cols, function(c) cor(df[[c]], df[[t]]))
  tibble(target = t, dim = ae_cols, r = rs)
})
ggplot(corr_long, aes(reorder(dim, r), r, fill = r > 0)) +
  geom_col() +
  facet_wrap(~ target, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("#c0392b", "#2980b9"), guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 5)) +
  labs(x = NULL, y = "Pearson r")
```

## Per-target spatial-CV RF

```{r multi-rfs}
rf_per_target <- map_dfr(target_names, function(t) {
  cv <- rf_spatial_cv(df, t, ae_cols)
  ci <- boot_mean_r2(cv$r2)
  tibble(target = t, mean_r2 = ci$mean, lo = ci$lo, hi = ci$hi,
         mean_rmse = mean(cv$rmse))
})
rf_per_target
```

```{r multi-pred-obs, fig.cap="Out-of-fold predicted vs.\\ observed for each target, ordered by mean R^2.", fig.height=6}
oof_per_target <- map_dfr(target_names, function(t) {
  cv <- rf_spatial_cv(df, t, ae_cols)
  tibble(target = t,
         pred = unlist(cv$pred),
         obs  = unlist(cv$obs))
})
ord <- rf_per_target %>% arrange(desc(mean_r2)) %>% pull(target)
oof_per_target$target <- factor(oof_per_target$target, levels = ord)
ggplot(oof_per_target, aes(obs, pred)) +
  geom_point(alpha = 0.25, size = 0.5, color = "#5b8def") +
  geom_abline(slope = 1, intercept = 0, color = "#c0392b", linetype = "dashed") +
  facet_wrap(~ target, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(x = "Observed", y = "Predicted")
```

# Phase C — Sentinel-2 baseline (Direction 2)

```{r load-s2}
s2 <- read.csv("../data/s2_features.csv")
df2 <- df %>% inner_join(s2, by = c("lon","lat"))
s2_features <- setdiff(colnames(s2), c("lon","lat"))
cat("S2 features:", length(s2_features), "  joined rows:", nrow(df2), "\n")

# Drop rows with any S2 NA (BSC-empty over evergreen forests etc.)
df2_full <- df2
df2 <- df2 %>% drop_na(all_of(s2_features))
cat("after dropping NA:", nrow(df2), " (lost", nrow(df2_full) - nrow(df2), ")\n")
df2 <- make_spatial_folds(df2)
```

## Univariate r comparison (top features per source)

```{r s2-corrs}
ae_r <- sapply(ae_cols, function(c) cor(df2[[c]], df2$log_soc))
s2_r <- sapply(s2_features, function(c) cor(df2[[c]], df2$log_soc))
top_compare <- bind_rows(
  tibble(source = "AE", feature = names(ae_r), r = ae_r),
  tibble(source = "S2", feature = names(s2_r), r = s2_r)
) %>% group_by(source) %>% slice_max(abs(r), n = 10) %>% ungroup()
top_compare
```

## Three-way RF benchmark

```{r s2-rfs}
rf_s2   <- rf_spatial_cv(df2, "log_soc", s2_features)
rf_ae   <- rf_spatial_cv(df2, "log_soc", ae_cols)
rf_both <- rf_spatial_cv(df2, "log_soc", c(s2_features, ae_cols))

bench <- tibble(
  model = c("S2 only","AE only","S2 + AE"),
  r2_mean = c(mean(rf_s2$r2), mean(rf_ae$r2), mean(rf_both$r2)),
  rmse    = c(mean(rf_s2$rmse), mean(rf_ae$rmse), mean(rf_both$rmse))
)
bench
```

## Paired bootstrap on $\Delta R^2$

```{r s2-paired-boot}
oof <- tibble(
  fold = unlist(rf_ae$fold),
  pred_ae = unlist(rf_ae$pred), pred_s2 = unlist(rf_s2$pred),
  pred_both = unlist(rf_both$pred), obs = unlist(rf_ae$obs)
)
# point-level squared errors
oof <- oof %>% mutate(
  se_ae = (pred_ae - obs)^2,
  se_s2 = (pred_s2 - obs)^2,
  se_both = (pred_both - obs)^2
)
# Per-fold R^2 -> bootstrap on the difference
fold_r2 <- oof %>% group_by(fold) %>%
  summarise(r2_ae = cor(pred_ae, obs)^2,
            r2_s2 = cor(pred_s2, obs)^2,
            r2_both = cor(pred_both, obs)^2,
            .groups = "drop")

paired_boot <- function(a, b, B = 10000) {
  d <- a - b
  bo <- boot::boot(d, function(x, idx) mean(x[idx]), R = B)
  ci <- boot::boot.ci(bo, type = "bca")$bca[4:5]
  list(delta = mean(d), lo = ci[1], hi = ci[2])
}
delta_ae_s2   <- paired_boot(fold_r2$r2_ae, fold_r2$r2_s2)
delta_both_ae <- paired_boot(fold_r2$r2_both, fold_r2$r2_ae)
delta_ae_s2; delta_both_ae
```

```{r s2-bar, fig.cap="Random-forest log-SOC R^2 across feature stacks. Error bars from per-fold bootstrap."}
plot_df <- tibble(
  model = factor(c("S2 only","AE only","S2 + AE"), levels = c("S2 only","AE only","S2 + AE")),
  r2 = c(mean(rf_s2$r2), mean(rf_ae$r2), mean(rf_both$r2)),
  lo = c(boot_mean_r2(rf_s2$r2)$lo, boot_mean_r2(rf_ae$r2)$lo, boot_mean_r2(rf_both$r2)$lo),
  hi = c(boot_mean_r2(rf_s2$r2)$hi, boot_mean_r2(rf_ae$r2)$hi, boot_mean_r2(rf_both$r2)$hi)
)
ggplot(plot_df, aes(model, r2, fill = model)) +
  geom_col() +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2) +
  scale_fill_manual(values = c("#27ae60","#5b8def","#34495e"), guide = "none") +
  theme_minimal() +
  labs(x = NULL, y = expression(R^2), title = "S2 vs. AlphaEarth vs. combined")
```

# Phase F — KSSL validation (Direction 3)

```{r kssl-load, eval=file.exists("../data/kssl_with_features.csv")}
kssl <- read.csv("../data/kssl_with_features.csv")
cat("KSSL rows:", nrow(kssl), "\n")
kssl$log_soc <- log1p(kssl$soc_pct_0_30 * 10)  # if in %, convert to g/kg-equivalent
kssl <- make_spatial_folds(kssl)

# OpenLandMap-vs-KSSL agreement
if ("openlandmap_soc" %in% colnames(kssl)) {
  agree <- cor(kssl$openlandmap_soc, kssl$soc_pct_0_30)^2
  cat("OpenLandMap-vs-KSSL R^2:", round(agree, 3), "\n")
}

rf_ae_kssl   <- rf_spatial_cv(kssl, "log_soc", ae_cols)
s2_kssl_features <- intersect(colnames(kssl), s2_features)
rf_s2_kssl   <- rf_spatial_cv(kssl, "log_soc", s2_kssl_features)
rf_both_kssl <- rf_spatial_cv(kssl, "log_soc", c(s2_kssl_features, ae_cols))

kssl_bench <- tibble(
  model = c("S2 only","AE only","S2 + AE"),
  r2 = c(mean(rf_s2_kssl$r2), mean(rf_ae_kssl$r2), mean(rf_both_kssl$r2)),
  rmse = c(mean(rf_s2_kssl$rmse), mean(rf_ae_kssl$rmse), mean(rf_both_kssl$rmse))
)
kssl_bench
```

## Moran's I on KSSL residuals

```{r morans-i, eval=file.exists("../data/kssl_with_features.csv")}
suppressPackageStartupMessages(library(spdep))
oof_resid <- tibble(
  lon = unlist(rf_ae_kssl$lon),
  lat = unlist(rf_ae_kssl$lat),
  resid = unlist(rf_ae_kssl$pred) - unlist(rf_ae_kssl$obs)
) %>% sample_n(min(2000, n()))  # cap for spdep memory

coords <- as.matrix(oof_resid[, c("lon","lat")])
nb <- knn2nb(knearneigh(coords, k = 8))
lw <- nb2listw(nb, style = "W")
mi <- moran.test(oof_resid$resid, lw, randomisation = TRUE)
print(mi)
```

# Headline comparison table

```{r headline}
headline <- bench %>%
  rename(`OpenLandMap R2` = r2_mean, `OpenLandMap RMSE` = rmse) %>%
  mutate(`KSSL R2` = if (exists("kssl_bench")) kssl_bench$r2[match(model, kssl_bench$model)] else NA_real_,
         `KSSL RMSE` = if (exists("kssl_bench")) kssl_bench$rmse[match(model, kssl_bench$model)] else NA_real_)
headline
```

```{r sessioninfo}
sessionInfo()
```
