---
title: "Predicting Soil Organic Carbon from AlphaEarth Embeddings"
subtitle: "STAT 4000/5000 Final Project --- Code Notebook"
author: "Luke Robinson"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 4, dpi = 300,
  fig.path = "figures/", dev = "png"
)
set.seed(42)
```

# Setup

```{r libs}
library(tidyverse)
library(boot)
library(corrplot)
library(car)
library(ranger)

dir.create("figures", showWarnings = FALSE)

df <- read.csv("soil_samples.csv")
ae_cols <- sprintf("A%02d", 0:63)
cat("Rows:", nrow(df), " Cols:", ncol(df), "\n")
```

`soil_samples.csv` was extracted from Google Earth Engine (`01_extract_data.py`).
Each row is a random point in the contiguous US with 64 AlphaEarth Foundations
2024 annual embedding bands (`A00`..`A63`), OpenLandMap Soil Organic Carbon at
0 cm (`soc`, g/kg), MODIS IGBP land cover (`land_cover`), and lon/lat.

```{r preprocess}
# IGBP land-cover labels
igbp_labels <- c(
  "1"  = "Evergreen Needleleaf Forest",
  "2"  = "Evergreen Broadleaf Forest",
  "3"  = "Deciduous Needleleaf Forest",
  "4"  = "Deciduous Broadleaf Forest",
  "5"  = "Mixed Forest",
  "6"  = "Closed Shrublands",
  "7"  = "Open Shrublands",
  "8"  = "Woody Savannas",
  "9"  = "Savannas",
  "10" = "Grasslands",
  "11" = "Permanent Wetlands",
  "12" = "Croplands",
  "13" = "Urban and Built-up",
  "14" = "Cropland/Natural Vegetation Mosaic",
  "15" = "Permanent Snow and Ice",
  "16" = "Barren",
  "17" = "Water Bodies"
)

df <- df %>%
  mutate(
    land_cover = as.character(land_cover),
    lc_label   = igbp_labels[land_cover],
    log_soc    = log1p(soc)
  )

# basic numeric checks
stopifnot(all(!is.na(df$soc)))
range(df$soc); median(df$soc)
```

# Method 1: Exploratory Data Analysis

We start with **exploratory data analysis** (numerical and graphical summaries
of the response variable, the predictors, and their bivariate relationships).

## Distribution of SOC

```{r eda-soc-hist, fig.cap="Distribution of soil organic carbon. Raw values are right-skewed; the log1p transform is approximately symmetric."}
p1 <- ggplot(df, aes(soc)) +
  geom_histogram(bins = 40, fill = "#5b8def", color = "white") +
  labs(x = "SOC (g/kg)", y = "Count", title = "Raw SOC") +
  theme_minimal()

p2 <- ggplot(df, aes(log_soc)) +
  geom_histogram(bins = 40, fill = "#27ae60", color = "white") +
  labs(x = "log(1 + SOC)", y = "Count", title = "log-transformed SOC") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

```{r eda-summary}
soc_summary <- df %>%
  summarise(
    n        = n(),
    mean     = mean(soc),
    sd       = sd(soc),
    median   = median(soc),
    iqr      = IQR(soc),
    skewness = mean((soc - mean(soc))^3) / sd(soc)^3
  )
soc_summary
```

The right-skew (`r round(soc_summary$skewness, 2)`) violates approximate
normality. We use `log1p(SOC)` for downstream inferential analyses.

## SOC by land cover

```{r eda-boxplot, fig.cap="log-SOC by IGBP land-cover class. Forest classes have the highest median; barren and shrubland the lowest."}
top_lc <- df %>% count(lc_label, sort = TRUE) %>% slice_head(n = 8) %>% pull(lc_label)

df %>%
  filter(lc_label %in% top_lc) %>%
  mutate(lc_label = fct_reorder(lc_label, log_soc, median)) %>%
  ggplot(aes(lc_label, log_soc, fill = lc_label)) +
  geom_boxplot(outlier.alpha = 0.4) +
  coord_flip() +
  labs(x = NULL, y = "log(1 + SOC)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Correlation between AlphaEarth dimensions and log-SOC

We examine the marginal Pearson correlation between each of the 64 embedding
dimensions and `log(1 + SOC)`.

```{r eda-corr-bar, fig.cap="Pearson correlation between each of the 64 AlphaEarth dimensions and log-SOC. Several dimensions exceed |r| = 0.3."}
cors <- sapply(ae_cols, function(c) cor(df[[c]], df$log_soc))
cor_df <- tibble(dim = ae_cols, r = cors) %>%
  arrange(desc(abs(r)))

ggplot(cor_df, aes(reorder(dim, r), r, fill = r > 0)) +
  geom_col() +
  scale_fill_manual(values = c("#c0392b", "#2980b9"), guide = "none") +
  labs(x = NULL, y = "Pearson r vs. log-SOC") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, size = 5))

top_dim <- cor_df$dim[1]
top_r   <- cor_df$r[1]
cat("Top dimension:", top_dim, "  r =", round(top_r, 3), "\n")
```

```{r eda-scatter, fig.cap="Scatter of log-SOC against the four AlphaEarth dimensions with the largest |r|."}
top4 <- cor_df$dim[1:4]
df_long <- df %>%
  pivot_longer(all_of(top4), names_to = "dim", values_to = "value")

ggplot(df_long, aes(value, log_soc)) +
  geom_point(alpha = 0.25, size = 0.6, color = "#34495e") +
  geom_smooth(method = "loess", se = FALSE, color = "#e67e22") +
  facet_wrap(~ dim, scales = "free_x") +
  labs(x = "AlphaEarth value", y = "log(1 + SOC)") +
  theme_minimal()
```

# Method 2: Hypothesis Testing

We test whether mean log-SOC differs between the two largest IGBP groups in
our sample, **Grasslands** (class 10) and **Croplands** (class 12). A
two-sample Welch's *t*-test is appropriate because (a) the groups are
independent, (b) the response is approximately symmetric on the log scale,
and (c) we do not assume equal variances.

```{r ht-setup}
two_groups <- df %>% filter(land_cover %in% c("10", "12"))
tab <- two_groups %>% count(lc_label)
tab
```

## Assumption checks

```{r ht-assumptions, fig.cap="QQ plots of log-SOC for the two groups. Both are close to linear, supporting an approximately normal sampling distribution of the mean (large n further by the CLT)."}
par(mfrow = c(1, 2))
qqnorm(two_groups$log_soc[two_groups$land_cover == "10"],
       main = "Grasslands"); qqline(two_groups$log_soc[two_groups$land_cover == "10"])
qqnorm(two_groups$log_soc[two_groups$land_cover == "12"],
       main = "Croplands"); qqline(two_groups$log_soc[two_groups$land_cover == "12"])
par(mfrow = c(1, 1))

# Levene's test for equal variances
levene_res <- car::leveneTest(log_soc ~ factor(land_cover), data = two_groups)
levene_res
```

Levene's test gives *p* = `r signif(levene_res[["Pr(>F)"]][1], 3)`; with
unequal variances likely, Welch's *t* is the right choice.

## Welch's two-sample *t*-test

```{r ht-ttest}
t_res <- t.test(log_soc ~ land_cover, data = two_groups,
                var.equal = FALSE, conf.level = 0.95)
t_res

# Cohen's d (using pooled sd for effect size)
g1 <- two_groups$log_soc[two_groups$land_cover == "10"]
g2 <- two_groups$log_soc[two_groups$land_cover == "12"]
pooled_sd <- sqrt(((length(g1)-1)*var(g1) + (length(g2)-1)*var(g2)) /
                  (length(g1)+length(g2)-2))
cohens_d <- (mean(g1) - mean(g2)) / pooled_sd
cat("Cohen's d =", round(cohens_d, 3), "\n")
```

# Method 3: Bootstrap Confidence Intervals

We use the **nonparametric bootstrap** (B = 10,000 resamples, BCa intervals)
to obtain confidence intervals for two quantities without distributional
assumptions: (i) the Pearson correlation between the strongest AlphaEarth
dimension and log-SOC, and (ii) the same group mean difference tested above.

## Bootstrap CI for the strongest correlation

```{r boot-cor}
B <- 10000
top_vec <- df[[top_dim]]
y_vec   <- df$log_soc

corr_stat <- function(d, idx) cor(d[idx, 1], d[idx, 2])

set.seed(42)
boot_cor <- boot(data.frame(x = top_vec, y = y_vec), corr_stat, R = B)
boot_cor_ci <- boot.ci(boot_cor, type = "bca")
boot_cor
boot_cor_ci
```

```{r boot-cor-fig, fig.cap="Bootstrap distribution (B=10,000) of the Pearson correlation between the strongest AlphaEarth dimension and log-SOC. The BCa 95\\% CI is shown as red dashed lines."}
ci <- boot_cor_ci$bca[4:5]
ggplot(data.frame(r = boot_cor$t[, 1]), aes(r)) +
  geom_histogram(bins = 60, fill = "#5b8def", color = "white") +
  geom_vline(xintercept = boot_cor$t0, color = "black", linewidth = 1) +
  geom_vline(xintercept = ci, color = "#c0392b", linetype = "dashed") +
  labs(x = paste0("Bootstrap r (", top_dim, " vs. log-SOC)"),
       y = "Frequency") +
  theme_minimal()
```

## Bootstrap CI for the group mean difference

```{r boot-meandiff}
mean_diff <- function(d, idx) {
  s <- d[idx, ]
  mean(s$log_soc[s$land_cover == "10"]) -
    mean(s$log_soc[s$land_cover == "12"])
}

set.seed(42)
boot_md <- boot(two_groups, mean_diff, R = B,
                strata = factor(two_groups$land_cover))
boot_md_ci <- boot.ci(boot_md, type = "bca")
boot_md
boot_md_ci

cat("\nFor comparison, Welch's t-test 95% CI:",
    signif(t_res$conf.int[1], 3), "to", signif(t_res$conf.int[2], 3), "\n")
```

```{r boot-md-fig, fig.cap="Bootstrap distribution (B=10,000) of the Grasslands $-$ Croplands log-SOC mean difference, with stratified resampling. BCa 95\\% CI in red."}
ci_md <- boot_md_ci$bca[4:5]
ggplot(data.frame(d = boot_md$t[, 1]), aes(d)) +
  geom_histogram(bins = 60, fill = "#27ae60", color = "white") +
  geom_vline(xintercept = boot_md$t0, color = "black", linewidth = 1) +
  geom_vline(xintercept = ci_md, color = "#c0392b", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "grey50", linetype = "dotted") +
  labs(x = "Bootstrap mean difference (Grasslands - Croplands)", y = "Frequency") +
  theme_minimal()
```

# Beyond the three: PCA and Random Forest

These analyses are *outside* the three class techniques but directly answer
the proposal's research question --- can the AlphaEarth latent space predict
SOC? --- and are reported as a supplementary analysis.

## Principal component analysis

```{r pca, fig.cap="Cumulative variance explained by AlphaEarth principal components."}
pca <- prcomp(df[, ae_cols], center = TRUE, scale. = FALSE)
varexp <- pca$sdev^2 / sum(pca$sdev^2)
cum    <- cumsum(varexp)

scree <- tibble(component = seq_along(varexp), cum_var = cum)
ggplot(scree, aes(component, cum_var)) +
  geom_line(color = "#5b8def", linewidth = 1) +
  geom_point(color = "#5b8def") +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "grey50") +
  labs(x = "Principal component", y = "Cumulative variance explained") +
  theme_minimal()

cat("PCs to explain 90% variance:", which(cum >= 0.90)[1], "\n")
cat("PCs to explain 95% variance:", which(cum >= 0.95)[1], "\n")
```

## Spatial-block cross-validated Random Forest

We assess predictive skill with 5-fold spatial cross-validation, splitting on
1° latitude/longitude blocks so neighbouring points cannot leak between train
and test.

```{r rf-cv}
df$block <- paste0(round(df$lon), "_", round(df$lat))
unique_blocks <- unique(df$block)
set.seed(42)
fold_assign <- sample(seq_len(5), length(unique_blocks), replace = TRUE)
df$fold <- fold_assign[match(df$block, unique_blocks)]

rf_form <- as.formula(
  paste("log_soc ~", paste(ae_cols, collapse = " + "))
)

cv_res <- map_dfr(1:5, function(k) {
  train <- df %>% filter(fold != k)
  test  <- df %>% filter(fold == k)
  m <- ranger(rf_form, data = train, num.trees = 500, seed = 42)
  pred <- predict(m, data = test)$predictions
  tibble(
    fold = k,
    n_test = nrow(test),
    rmse = sqrt(mean((pred - test$log_soc)^2)),
    r2   = cor(pred, test$log_soc)^2
  )
})
cv_res
cat("Mean R^2:", round(mean(cv_res$r2), 3),
    "  Mean RMSE:", round(mean(cv_res$rmse), 3), "\n")
```

```{r rf-importance, fig.cap="Top 15 AlphaEarth dimensions by Random Forest variable importance (impurity)."}
m_full <- ranger(rf_form, data = df, num.trees = 500,
                 importance = "impurity", seed = 42)
imp <- enframe(m_full$variable.importance, "dim", "importance") %>%
  arrange(desc(importance)) %>% slice_head(n = 15)

ggplot(imp, aes(reorder(dim, importance), importance)) +
  geom_col(fill = "#34495e") +
  coord_flip() +
  labs(x = NULL, y = "Random Forest impurity importance") +
  theme_minimal()
```

```{r rf-pred-fig, fig.cap="Out-of-fold predicted vs.~observed log-SOC across the 5 spatial folds."}
oof <- map_dfr(1:5, function(k) {
  train <- df %>% filter(fold != k)
  test  <- df %>% filter(fold == k)
  m <- ranger(rf_form, data = train, num.trees = 500, seed = 42)
  test$pred <- predict(m, data = test)$predictions
  test
})

ggplot(oof, aes(log_soc, pred)) +
  geom_point(alpha = 0.3, size = 0.7, color = "#5b8def") +
  geom_abline(slope = 1, intercept = 0, color = "#c0392b", linetype = "dashed") +
  labs(x = "Observed log(1 + SOC)", y = "Predicted log(1 + SOC)") +
  theme_minimal()
```

# Session info

```{r sessioninfo}
sessionInfo()
```
